A Systematic Literature Review on Extensions to Role-Based Access Control

### 2013-05-12 paper/sections/abstract.tex
Updates to address dr. williams' comments

\noindent
{\bf Context}: Since the National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) standard using RBAC Reference Model in the late 1990's, computing in new domains has led to proposals for extensions to the RBAC Reference Model; for example adding context around permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how RBAC extension models are evaluated by providing an assessment of the state of the RBAC extension models. We accomplished this through
establishment of a set of extension categories, an examination of the state of the art in evaluations of the RBAC extension models, and a breakdown of the motivations that have led to the RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as the RBAC extension models. 
\\
\\
\noindent
{\bf Results}: Our results show that the RBAC extension models can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. We identified only
8 of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with 9 of the 28 models, enterprise workflows with 5 of 28, and mobile computing with 5 of 28.
\\
\\
\noindent
{\bf Conclusions}: When examined as an entire group, the eight categories can be aligned under the single category of context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with 1
paper doing an evaluation comparing itself to the RBAC Reference Model, 8 presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The landscape of the RBAC extension models serves as a basis from which a new, and vetted, RBAC extension model should be derived.

>>>>

\noindent
{\bf Context}: Since the US National Institute of Standards and Technology (NIST) proposed Role-Based Access Control (RBAC) as documented in the RBAC Reference Model in the late 1990's, domain-specific extensions have been proposed.  For example, the mobile domain has surfaced a need for permission granting and role activation.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how RBAC extension models are evaluated by providing an assessment of the state of the RBAC extension models.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that began with 1,716 paper of which 28 were deemed primary sources for inclusion as the RBAC extension models. We established a set of extension categories, examined the state of the art in evaluations of the RBAC extension models, and categorized the motivations that have led to the RBAC extension models.
\\
\\
\noindent
{\bf Results}: We classified the RBAC extension models into eight extension categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial, and Temporal. only 8 of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. The primary domains 
that inspired extensions were the medical domain with 9 of the 28 models, enterprise workflows with 5 of 28, and mobile computing with 5 of 28.
\\
\\
\noindent
{\bf Conclusions}: All of the eight RBAC extension categories deal with context. 
Our literature review shows that the state of evaluation of RBAC model evaluation is limited with 1
paper doing an evaluation comparing itself to the RBAC Reference Model, 8 presenting example 
scenarios of their model in action, and 12 of the 28 models lacking an evaluation that we could discern. The magnitude of extensions to the RBAC standard suggests a revised standard may be beneficial.

### 2013-05-12 paper/sections/core_rbac.tex
Updates to address dr. williams' comments

NIST's work can directly benefit organizations by lowering cost of early phase Research and Development, and implementation of RBAC.
Since the RBAC standard was first introduced, a 2010 report by RTI International showed that the rate of RBAC adoption has rapidly grown over recent years \cite{o20102010}. 
The analysts estimate that by the end of 2010 at least a portion of permission of systems will use RBAC for more than half of users at organizations with more than 500 employees. The analysts estimate that RBAC technology has generated \$6.1 billion in net economic benefits to industry where NIST RBAC standard saved \$1.1 billion.

The NIST RBAC standard proposed by Ferraiolo et al. \cite{ferraiolo} and later adopted as the official standard for RBAC by the INCITS includes three components of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC. Each component includes its corresponding RBAC Reference model. We first describe the core RBAC, associated entities and other terminology encountered across the space of our review. We next describe hierarchical RBAC and constrained RBAC, which are developed by incorporating new features into the core RBAC.

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item a set of \emph{Users}: A user can be a person or an agent.
\item a set of \emph{Roles}: A role is a collection of permissions to perform a specific job function in an organization.
\item a set of \emph{Permissions}: A permission refers to an access mode that can be exercised on an object in the system and a session relates a user to possibly many roles.
\item a set of \emph{Sessions}: In each session, a user can be assigned to some of the roles, only when the corresponding role is enabled for activation for that time.		
\end{itemize}

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic entities, two functions are defined:
user assignment ($UA$) and permission assignment ($PA$) functions.
$UA$ represents assignment of users to roles.
$PA$ represents assignment of permissions to roles.

\begin{figure}[ht]
    \centering
        \includegraphics[width=4.0in]{sections/core-model.png}
    \caption{\label{fig:overview}Diagram in Core RBAC reference model\cite{ferraiolokuhn}.}
\end{figure}

@@

The hierarchical RBAC Reference Model adds role hierarchies ($RH$) feature to the core RBAC Reference Model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of an employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. Formally, the role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing the many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one role.

@@

\begin{itemize}
	\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-roles, which can be assigned to a single user statically. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. On situations
	where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ statically. SSoD is known to be too rigid for practical use in cases where a user should have permissions when a user is assigned to both $Role_A$ and $Role_B$.
	\item Dynamic Separation of Duty (DSD): Dynamic SoD is known to be
more flexible than SSoD. DSD restricts the conflicting-role assignments dynamically that are associated with a user. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. For the situations where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ dynamically.	
\end{itemize}

>>>>

NIST's work can directly benefit organizations by lowering cost of early phase Research and Development, and implementation of RBAC.
Since the RBAC standard was first introduced, a 2010 report by RTI International showed that the rate of RBAC adoption has rapidly grown~\cite{o20102010}. 
The analysts estimate that RBAC technology has generated \$6.1 billion in net economic benefits to industry where NIST RBAC standard saved \$1.1 billion.

The RBAC standard includes three components of RBAC: core RBAC, hierarchical RBAC, and constrained RBAC. Each component includes its corresponding RBAC Reference model. We first describe the core RBAC, associated entities and other terminology. We next describe hierarchical RBAC and constrained RBAC, which are developed by incorporating new features into the core RBAC.

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item \emph{Users}: A user is defined as a human being. Although the concept of a user can be extended to include machines, networks, or intelligent autonomous agents, the definition is limited to a person in the RBAC standard. 
\item \emph{Roles}: A role is a job function within the context of an organization with some associated semantics regarding the authority and responsibility conferred on the user assigned to the role.
\item \emph{Permissions}: A permission is an approval to perform an operation on one or more RBAC protected objects in the system.
\item \emph{Sessions}: A session is a mapping between a user and an activated subset of roles that are assigned to the user.
\end{itemize}

In RBAC, a user can exercise a permission only if the user is assigned to their role.
In addition to the four basic entities, two functions are defined:
user assignment ($UA$) and permission assignment ($PA$) functions.
$UA$ represents assignment of users to roles.
$PA$ represents assignment of permissions to roles.

\begin{figure}[ht]
    \centering
        \includegraphics[width=4.0in]{sections/core-model.eps}
    \caption{\label{fig:overview}Diagram in Core RBAC reference model\cite{ferraiolokuhn}.}
\end{figure}

@@

The hierarchical RBAC Reference Model adds role hierarchies ($RH$) feature to the core RBAC Reference Model.
$RH$ incorporates a structure of roles in an organization using inheritance relationships among attributes such as roles.
The role structure in an organization may use
a role $r_1$, which inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of an employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments for a user. 
Concept of the role inheritance describes the many-to-many mapping role inheritance relations among roles.
Therefore, more than one role (e.g., two roles $r_1$ and $r_1'$) can inherit all permissions of $r_2$ 
General role hierarchies can be extended to use the concept of multiple inheritances where
$r_1$ inherits all permissions from more than one role (e.g., two roles $r_2$ and $r_2'$) .

@@

\begin{itemize}
	\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-roles, which can be assigned to a single user statically. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. On situations
	where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ statically. SSoD is known to be too rigid for practical use in cases where a user should have permissions when a user is assigned to both $Role_A$ and $Role_B$.
	For example, accounting clerk role requests a check and accounting manager role approves the check. These two roles must be
	mutually exclusive to avoid a situation where one approves the check that the one requested.

	\item Dynamic Separation of Duty (DSD): Dynamic SoD is known to be
more flexible than SSoD. DSD restricts the conflicting-role assignments dynamically that are associated with a user. Consider that roles $Role_A$ and $Role_B$ are conflicting with each other. For the situations where multiple roles can be associated with a single user, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$ dynamically.	
Consider that one can be assigned to the accounting clerk and the accounting manager role at the same time.
In such a situation, DSD can enforce that the accounting manager role does not approve her/his requested checks but
approve only checks that others requested. This feature requires handling dynamic environments.
\end{itemize}

### 2013-05-12 paper/sections/intro.tex
Updates to address dr. williams' comments

Role-Based Access Control (RBAC) is widely used for maintaining and managing an organization's 
access control based on assigning permissions to roles, and roles to users, instead of 
assigning the permissions directly to individual users.
RBAC is used for securing 
various applications including web services, database applications, and healthcare applications. 
In RBAC, roles represent a set of permissions needed to perform a particular job function within an organization.  
Multiple users who are involved in that specific job function within the organization can then be 
assigned to a single role to inherit the required access. 
The ability to logically group users into roles associated with permissions becomes paramount for managing access control as an organization grows
and the number of permissions and users scales upward.
As permissions can be managed by a role instead of a user, RBAC has an advantage to significantly reduce complexity of security
administration.
For example, if a user requires access to resources 
associated with a manager role within a given organization, security policy administrators need only associate the user with the manager role instead
of assigning permissions of the manager role to the user.

The use of RBAC has become popular since the National Institute for Standards and Technology (NIST) 
first proposed RBAC standard in the 1990s \cite{o20102010}.
The RBAC standard was first proposed when NIST 
requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework 
proposed by Sandhu et al.~\cite{sandhu1996role}.  
In 2004, this standard was adopted as ANSI/INCITS 359-2004 approved by American National Standards Institute (ANSI) and the InterNational Committee for Information Technology Standards (INCITS). 
The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2010\footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}} that showed the cost savings of RBAC implementation and maintenance. 
Prior to the development of the RBAC standard, vendors proposed and implemented their own defined RBAC features without general agreement on a unified definition of RBAC or feature set. The RBAC standard includes RBAC Reference Model. The RBAC Reference Model serves as a basis for defining scope of RBAC features and functional specifications of RBAC features.

Innovation and the use of RBAC in new domains have led to scenarios that, by some accounts, the RBAC Reference Model cannot handle \cite{kuhn2010adding}.  
For example, Ni et al. \cite{ni2010privacy} state that the RBAC Reference Model is ``not designed to enforce privacy policies and barely 
meet privacy protection requirements'' with the introduction of privacy concerns in domains such as healthcare. 
Since the introduction of the RBAC standard, RBAC extension models have appeared in the literature, each adding one or more features on top of components in the RBAC Reference Model.
%A cursory examination of the extension model papers does not provide good comparisons for developers and researchers to build upon. 
Further, these extension models are each building upon and adding features to a standard that was designed to reduce the economic impact experienced by enterprises and increase interoperability \cite{o20102010}.

\textit{The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how the RBAC extension models are evaluated by providing an assessment of RBAC extensions. We accomplished this through 
establishment of a set of RBAC extension categories, an examination of the state of the art in evaluations of the RBAC extension models, and a breakdown of the motivations that have led to the RBAC extension models.} To accomplish this goal, we seek to answer following research questions:

@@

We performed a systematic literature review designed to explore the current body of research in the area of extensions to the RBAC Reference Model. 
The review we performed yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as extension models to the RBAC Reference Model.  
This review is intended to serve as a starting place for researchers looking to tackle new problems in the realm of authorization and as a reference guide for discovering
the current state of the RBAC extension models. 
For developers looking to find a model to fit their potentially unique access control needs, this review provides a basis for comparison and easy look up for what extension model to use.  
Furthermore, the review provides insight into the state of the art in evaluating RBAC extension models.

Our research provides contributions to the community in following ways:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Summarizes current research on the RBAC extension models.
\item Guides a direction for a new standard based on the RBAC extension models.
\end{itemize}

>>>>

Software systems use access control mechanisms to determine which subjects can access
which resources.
Role-Based Access Control (RBAC) is a widely used access control mechanism designed for maintaining and managing an organization's 
access control based on assigning permissions to roles, and roles to users, instead of 
assigning the permissions directly to individual users.
RBAC is used for securing 
various applications including web services, database applications, and healthcare applications. 
In RBAC, roles represent a set of permissions needed to perform a particular job function within an organization.  
Multiple users who are involved in that specific job function within the organization can then be 
assigned to a single role to inherit the required access. 
The ability to logically group users into roles associated with permissions becomes paramount for managing access control as an organization grows
and the number of permissions and users scales upward.
As permissions can be managed by a role instead of a user, RBAC has an advantage to significantly reduce complexity of security
administration.
For example, if a user requires access to resources 
associated with a manager role within a given organization, security policy administrators need only associate the user with the manager role instead
of assigning a set of individual permissions of the manager role.

The use of RBAC has become popular since the National Institute for Standards and Technology (NIST) 
first proposed the RBAC standard in the 1990s \cite{o20102010}.
The RBAC standard was first proposed when NIST 
requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework 
proposed by Sandhu et al.~\cite{sandhu1996role}.  
In 2004, this standard was adopted as ANSI/INCITS 359-2004 approved by American National Standards Institute (ANSI) and the InterNational Committee for Information Technology Standards (INCITS). 
The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2010\footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}} that showed the cost savings of RBAC implementation and maintenance. 
Prior to the development of the RBAC standard, vendors proposed and implemented their own defined RBAC without general agreement on a unified definition of RBAC or RBAC features (e.g., inheritance relationships among roles). The RBAC standard includes the RBAC Reference Model. The RBAC Reference Model serves as a basis for defining scope of RBAC features and functional specifications of RBAC features.

Since the introduction of the RBAC standard, researchers proposed domain-specific extensions such as features of dynamic context or privacy, which
can be critical in medical and mobile domains. RBAC extension models have appeared in the literature, each adding one or more features on top of components in the RBAC Reference Model \cite{kuhn2010adding}.
For example, Ni et al. \cite{ni2010privacy} state that the RBAC Reference Model is ``not designed to enforce privacy policies and barely 
meet privacy protection requirements'', with the introduction of privacy concerns in the medical domain.
They proposed an RBAC extension model for privacy concerns incorporated in the RBAC Reference Model.
%A cursory examination of the extension model papers does not provide good comparisons for developers and researchers to build upon. 
Further, these extension models are each building upon and adding features to a standard that was designed to reduce the economic impact experienced by enterprises and increase interoperability \cite{o20102010}.

\textit{The goal of our work is to aid practitioners and researchers in choosing an RBAC extension model, and in understanding
how the RBAC extension models are evaluated by providing an assessment of RBAC extensions. We established a set of extension categories, examined the state of the art in evaluations of the RBAC extension models, and categorized the motivations that have led to the RBAC extension models.
} To accomplish this goal, we seek to answer following research questions:

@@

We performed a systematic literature review to explore the current body of research in the area of extensions to the RBAC Reference Model. 
The review began with 1,716 papers, of which 28 were deemed primary sources for inclusion as extension models to the RBAC Reference Model.
%This review is intended to serve as a starting place for researchers looking to tackle new problems in the realm of authorization and as a reference guide for discovering
%the current state of the RBAC extension models. 
%For developers looking to find a model to fit their potentially unique access control needs, this review provides a basis for comparison and easy look up for what extension model to use.  
%Furthermore, the review provides insight into the state of the art in evaluating RBAC extension models.
Our research provides contributions to the community in following ways:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Serves as a starting place for researchers looking to tackle new problems in the realm of authorization and as a reference guide for discovering
the current state of the RBAC extension models. 
\item Provides a basis for comparison and easy look up for what extension model to use for developers looking to find a model to fit their potentially unique access control needs
\item Provides insight in evaluating RBAC extension models.
%\item Summarizes current research on the RBAC extension models.
%\item Guides a direction for a new standard based on the RBAC extension models.
\end{itemize}

### 2013-05-12 paper/sections/process.tex
Updates to address dr. williams' comments

The result set size stopping criteria was chosen due to a technical limitation of some search engines.  The stopping criteria related to the last ten titles are meant to stop after relevant results are no longer being returned by the search engine.  After gathering all 12 data sets, we combined the papers into a master list, which includes only distinct papers by systematically comparing the bibliographic information for each.
%The master list of 1,716 papers was used to perform a series of elimination rounds to narrow the list of papers and identify primary sources.
Out of the master list of 1,716 papers, two reviewers, namely Reviewer 1 and Reviewer 2, conducted a series of elimination rounds to narrow the list of papers and identify primary sources. Table \ref{tab:eliminations} shows the total number of papers selected by each reviewer for each round and how many papers from the disjoint set for each round survived to the next round.

@@

\multicolumn{1}{ |c| }{Reviewer 1}  &  & 305 & 86 & 46 \\ \hline
\multicolumn{1}{ |c| }{Reviewer 2} &  & 176 & 102 & 42 \\ \hline

>>>>

The result set size stopping criteria was chosen due to a technical limitation of some search engines.  The stopping criteria related to the last ten titles are meant to stop after relevant results are no longer being returned by the search engine.  After gathering all 12 data sets, we combined the papers into a master list, which includes only distinct papers by systematically comparing the bibliographic information for each.
%The master list of 1,716 papers was used to perform a series of elimination rounds to narrow the list of papers and identify primary sources.
Out of the master list of 1,716 papers, two reviewers, namely Author 1 and Author 2, conducted a series of elimination rounds to narrow the list of papers and identify primary sources. Table \ref{tab:eliminations} shows the total number of papers selected by each reviewer for each round and how many papers from the disjoint set for each round survived to the next round.

@@

\multicolumn{1}{ |c| }{Author 1}  &  & 305 & 86 & 46 \\ \hline
\multicolumn{1}{ |c| }{Author 2} &  & 176 & 102 & 42 \\ \hline
